% UPenn Master's Thesis - PIVOT
% Based on University of Pennsylvania Ph.D. Dissertation LaTeX Template v.20221216
% Modified for Master of Science in Engineering
%-------------------------------------------
\documentclass[12pt]{report}
\usepackage{upennstyle_master}

%-------------------------------------------
% Additional packages
\usepackage[nonamebreak,numbers,sort&compress]{natbib}
\bibliographystyle{ieeetr}
\usepackage[mathcal]{eucal}

\newcommand{\argminB}{\mathop{\mathrm{arg\,min}}}
%-------------------------------------------

%%%%% PRELIMINARY PAGES %%%%%
\title{PIVOT: PHYSICS INFORMED AND VISION BASED ONLINE TUNING}
\author{Haiyue Chen}
\gradgroup{Robotics}
\date{2026}
\supervisor{Supervisor Name}
\supervisortitle{Full Faculty Title, Department}
\gradchair{Graduate Group Chairperson Name, Full Faculty Title}
\committee{Committee Member 1, Full Faculty Title and Affiliation}
\committee{Committee Member 2, Full Faculty Title and Affiliation}

\authorlegal{Haiyue Chen}

%\dedication{Write your dedication text in italics here.}

\acknowledgement{Write your acknowledgement text here.}

\abstract{Accurate dynamics models are essential for Model Predictive Control (MPC) of autonomous vehicles, yet the relationship between model architecture and closed-loop control performance remains poorly characterized. Fixed-parameter physics models require manual calibration and degrade when operating conditions change, while purely data-driven models demand large datasets and are sensitive to real-world sensor noise. This thesis presents PIVOT (Physics Informed and Vision Based Online Tuning), a framework that addresses these limitations through two contributions. First, PIVOT introduces the Learned Single-Track model (LST), which augments a nonlinear bicycle model with state-dependent multiplicative gains and additive biases predicted by a neural network. The network receives DINOv2 visual features as auxiliary input, allowing the dynamics to adapt to the current operating environment without explicit surface classification. Second, PIVOT provides a systematic comparison of seven dynamics models spanning three categories (fixed-parameter physics, physics-informed learned, and purely data-driven), all trained on identical trajectory data from a 1/5th-scale autonomous off-road race car. The evaluation comprises two stages: offline trajectory prediction on held-out data and closed-loop racing using Safe Information-Theoretic Learning MPC (SIT-LMPC). On the offline prediction task, the LST model achieves a pose loss of 0.573, reducing error by 2.2$\times$ compared to the best data-driven baseline (TCN, 1.242) and by 56\% compared to the fixed-parameter bicycle model (DST, 1.302). An ablation over input features reveals that LST maintains a pose loss of 0.514 even without visual input, while the data-driven models degrade by 1.4--2.4$\times$ under the same condition. These results indicate that the nonlinear physics structure in LST provides a strong inductive bias that improves both accuracy and robustness relative to less constrained architectures on noisy real-world data.}

%-------------------------------------------
\begin{document}

%%% Title Page (page i, no number printed)
\maketitle
\setcounter{page}{2}

%%% Dedication (optional)
%\makededication

%%% Acknowledgment (optional)
\makeacknowledgement

%%% Abstract (optional)
\makeabstract

%%% Table of Contents (mandatory for >= 50 pages)
\tableofcontents

%%% List of Tables (optional)
\clearpage \phantomsection \addcontentsline{toc}{chapter}{LIST OF TABLES} \begin{singlespacing} \listoftables \end{singlespacing}

%%% List of Illustrations (optional)
%\clearpage \phantomsection \addcontentsline{toc}{chapter}{LIST OF ILLUSTRATIONS} \begin{singlespacing} \listoffigures \end{singlespacing}

%%%%% MAIN TEXT %%%%%
\begin{mainf}

\input{sections/introduction}

\input{sections/related_work}

\input{sections/problem}

\input{sections/method}

\input{sections/control}

\input{sections/experiments}

\input{sections/conclusion}

\end{mainf}

%%%%% BIBLIOGRAPHY %%%%%
\begin{bibliof}
\bibliography{bibliography}
\end{bibliof}

\end{document}
