\chapter{Model Architectures}
\label{sec:models}

This chapter presents seven modeling approaches of increasing expressiveness, all trained on the same trajectory dataset $\mathcal{D}_{\text{train}}$. This controlled setup isolates the effect of model architecture on prediction accuracy and control performance. The models fall into three categories: fixed-parameter physics models (Section~\ref{sec:fixed}), physics-informed learned models (Section~\ref{sec:physics-learned}), and purely data-driven models (Section~\ref{sec:data-driven}). Table~\ref{tab:models} summarizes the key characteristics of each architecture, and Table~\ref{tab:param-counts} reports the parameter counts.

All five learned models (LST, LTVL, MLP, LSTM, TCN) share a common two-stage architecture. A parameter network ($\approx$858K parameters) maps input features (DINOv2 embeddings and state context) to the dynamics model parameters. The dynamics model consumes these predicted parameters and rolls out over the 9-step prediction horizon. The parameter network and dynamics model are trained jointly: the optimizer backpropagates the yaw-normalized L1 loss through the integrator and dynamics equations into the parameter network weights, which are the only trainable variables. The dynamics computation is fully differentiable in JAX, and optimization employs Adam with a learning rate of $10^{-3}$.

The models differ in the number of parameters the network must predict (Table~\ref{tab:param-counts}). The LST model requires only 8 outputs (4 multiplicative gains and 4 additive biases), because the nonlinear bicycle dynamics handle the bulk of the prediction. The data-driven models require the parameter network to output the entire weight set of the dynamics model: 838 values for the MLP, 821 for the TCN, and 19,141 for the LSTM. Although these dynamics models are small by deep learning standards, the parameter network's output layer scales with the target dimensionality (e.g., $\approx$4.9M parameters for the LSTM output layer alone, 5.7$\times$ the shared backbone). This expansion amplifies the risk of overfitting on limited real-world data.

%--------------------------------------------------
\section{Fixed-Parameter Physics Models}
\label{sec:fixed}

These models employ conventional offline system identification with no learned components, serving as baselines for the comparison.

\subsection{Default Single-Track Model (DST)}
\label{sec:dst}

The default single-track model is a nonlinear bicycle model with a 7-dimensional state
\begin{equation}
  s = [x,\, y,\, \delta,\, v,\, \psi,\, \dot{\psi},\, \beta]^\top,
\end{equation}
representing position $(x, y)$, steering angle $\delta$, longitudinal velocity $v$, heading $\psi$, yaw rate $\dot{\psi}$, and sideslip angle $\beta$. The two control inputs are
\begin{equation}
  a = [\dot{\delta},\, a_{\text{long}}]^\top,
\end{equation}
where $\dot{\delta}$ denotes the steering rate and $a_{\text{long}}$ the longitudinal acceleration.

The model employs a velocity-dependent switching mechanism. At low speeds ($|v| < 1.5$\,m/s), a kinematic formulation governs the dynamics. At higher speeds, a dynamic formulation with nonlinear tire force modeling becomes active. The continuous-time dynamics are discretized via fourth-order Runge-Kutta (RK4) integration with time step $\Delta t$.

The 14 physical parameters, including mass ($m = 3.74$\,kg), yaw inertia ($I_z = 0.047$\,kg$\cdot$m$^2$), front and rear cornering stiffness ($C_{Sf} = 4.72$, $C_{Sr} = 5.46$), and tire-road friction coefficient ($\mu = 1.05$), are identified offline via system identification. State constraints enforce steering limits of $\pm 0.42$\,rad, velocity bounds of $[-5, 20]$\,m/s, and steering rate limits of $\pm 3.2$\,rad/s.

\subsection{Default Time-Varying Linear Model (DTVL)}
\label{sec:dtvl}

The default time-varying linear model adopts a 5-dimensional state $s = [x,\, y,\, \delta,\, v,\, \psi]^\top$ with the same two control inputs. The dynamics take the form
\begin{equation}
  \dot{s} = A \begin{bmatrix} s \\ a \end{bmatrix} + B,
  \label{eq:tvl}
\end{equation}
where $A \in \mathbb{R}^{5 \times 7}$ and $B \in \mathbb{R}^{5}$ are fixed matrices identified offline. This model captures dynamics only in the vicinity of the nominal operating point around which linearization is performed, limiting its efficacy under aggressive maneuvers.

%--------------------------------------------------
\section{Physics-Informed Learned Models (PIVOT)}
\label{sec:physics-learned}

The two proposed PIVOT models retain known physics structure while leveraging neural networks to adapt model parameters or system matrices as functions of the current state and action. By preserving established dynamics equations as an inductive bias, these architectures target improved sample efficiency and generalization compared to purely data-driven alternatives.

\subsection{Learned Single-Track Model (LST)}
\label{sec:lst}

The learned single-track model shares the same nonlinear bicycle dynamics and velocity-dependent switching mechanism as the DST model (Section~\ref{sec:dst}), but augments four dynamics channels with learnable corrections. Specifically, the steering rate ($\dot{\delta}$), longitudinal acceleration ($\dot{v}$), yaw acceleration ($\ddot{\psi}$), and sideslip rate ($\dot{\beta}$) channels are each modified as
\begin{equation}
  \dot{s}_i = f_{\text{physics},i}(s, a) \cdot \alpha_i + \beta_i,
  \label{eq:lst-augment}
\end{equation}
where $f_{\text{physics},i}$ denotes the original physics-based computation for the $i$-th channel, and $\alpha_i$, $\beta_i$ are learnable multiplicative gains and additive biases, respectively.

The shared parameter network predicts these 8 augmentation values from the current input features:
\begin{equation}
  [\alpha_1, \beta_1, \ldots, \alpha_4, \beta_4] = \mathcal{F}_{\text{LST}}(\text{features};\, \phi_{\text{LST}}).
\end{equation}
The gains and biases are initialized to $\alpha_i = 1$ and $\beta_i = 0$, ensuring that the model recovers the default single-track dynamics at initialization. The augmented dynamics are integrated via RK4, and the network is trained end-to-end by backpropagating through the integrator:
\begin{equation}
  \mathcal{L}_{\text{LST}} = \mathbb{E}_{(s,a,s') \in \mathcal{D}_{\text{train}}}\!\left[\left\|s' - \text{RK4}\big(\dot{s}(s, a;\, \phi_{\text{LST}}),\, \Delta t\big)\right\|^2\right].
\end{equation}

This formulation enables state-dependent correction of model mismatch, such as unmodeled tire saturation or surface friction variation, while retaining the known physics structure as an inductive bias.

\subsection{Learned Time-Varying Linear Model (LTVL)}
\label{sec:ltvl}

The learned time-varying linear model extends the DTVL model (Section~\ref{sec:dtvl}) by replacing the fixed matrices with state-dependent ones predicted by the shared parameter network:
\begin{equation}
  \dot{s} = A(s, a)\begin{bmatrix} s \\ a \end{bmatrix} + B(s, a),
  \label{eq:ltvl}
\end{equation}
where
\begin{equation}
  [A(s, a),\, B(s, a)] = \mathcal{F}_{\text{LTVL}}(\text{features};\, \phi_{\text{LTVL}}).
\end{equation}
The parameter network outputs the 40 values comprising the flattened matrices $A \in \mathbb{R}^{5 \times 7}$ and bias $B \in \mathbb{R}^{5}$. The predicted dynamics are integrated via RK4, and the network is trained to minimize one-step prediction error:
\begin{equation}
  \mathcal{L}_{\text{LTVL}} = \mathbb{E}_{(s,a,s') \in \mathcal{D}_{\text{train}}}\!\left[\left\|s' - \text{RK4}\big(\dot{s}(s, a;\, \phi_{\text{LTVL}}),\, \Delta t\big)\right\|^2\right].
\end{equation}

In contrast to the DTVL model, this formulation can represent globally nonlinear dynamics through state-dependent linearization while preserving linear structure at each operating point, offering a trade-off between the interpretability of linear models and the expressiveness of nonlinear ones.

%--------------------------------------------------
\section{Purely Data-Driven Models}
\label{sec:data-driven}

These models learn dynamics entirely from data without leveraging explicit physics structure. All three adopt the same 5-dimensional state $s = [x,\, y,\, \delta,\, v,\, \psi]^\top$ and two control inputs as the DTVL model, and predict state derivatives $\dot{s}$ with a time step of $\Delta t = 0.05$\,s.

\subsection{Multi-Layer Perceptron (MLP)}
\label{sec:mlp}

A feedforward network with a single hidden layer of 64 units and tanh activations maps the concatenated state-action vector to state derivatives:
\begin{equation}
  \dot{s} = W_2 \tanh\!\big(\alpha\,(W_1 [s^\top, a^\top]^\top + b_1)\big) + b_2,
\end{equation}
where $\alpha$ is a learnable activation scale. As a memoryless model, the MLP captures instantaneous nonlinear mappings but cannot exploit temporal structure in the data.

\subsection{Long Short-Term Memory Network (LSTM)}
\label{sec:lstm}

The LSTM employs a single-layer cell with 64 hidden units and four standard gates (input, forget, cell candidate, and output), allowing the network to selectively retain or discard information across timesteps. The heading angle $\psi$ is represented as $[\sin\psi,\, \cos\psi]$ to avoid the discontinuity at $\pm\pi$, yielding the input vector $[\tilde{s}_t^\top, a_t^\top]^\top \in \mathbb{R}^{8}$ with $\tilde{s}_t = [x,\, y,\, \delta,\, v,\, \sin\psi,\, \cos\psi]^\top$. The LSTM cell processes this input and produces a state update via an output projection:
\begin{equation}
  h_t, c_t = \text{LSTM}(\tilde{s}_t, a_t, h_{t-1}, c_{t-1};\, \phi_{\text{LSTM}}), \qquad \Delta s_t = W_{\text{out}}\, h_t + b_{\text{out}}.
\end{equation} Euler integration yields the next state as $s_{t+1} = s_t + \Delta s_t$, rather than the RK4 scheme employed by the physics-based models. RK4 evaluates the dynamics function at multiple intermediate points per timestep. For a recurrent model, each intermediate evaluation advances the hidden state, causing the memory to drift from the trajectory it is meant to represent. Euler integration evaluates the LSTM cell exactly once per timestep, preserving a consistent correspondence between hidden state and trajectory history.

During training, a burn-in phase (a sequence of preceding timesteps processed without loss computation) initializes the hidden state before the loss is evaluated. The training objective minimizes prediction error over multi-step autoregressive rollouts:
\begin{equation}
  \mathcal{L}_{\text{LSTM}} = \mathbb{E}_{\text{traj}}\!\left[\sum_{t=1}^{T_{\text{seq}}} \|s_t - \hat{s}_t\|^2\right],
\end{equation}
where $\hat{s}_t$ denotes the autoregressive prediction, encouraging the model to mitigate compounding errors over the prediction horizon.

\subsection{Temporal Convolutional Network (TCN)}
\label{sec:tcn}

The TCN maintains a sliding buffer of the $K = 4$ most recent state-action pairs. At each timestep, the current pair replaces the oldest entry, and the network processes the buffer through two causal 1D convolutional layers (kernel sizes 4 and 1, with a hidden dimension of 16 and tanh activations) followed by a linear output projection to state derivatives:
\begin{equation}
  \dot{s}_t = W_{\text{out}}\, \text{Conv}_2\!\big(\tanh(\text{Conv}_1(\text{buf}_t))\big) + b_{\text{out}}.
\end{equation}

Unlike the LSTM, the TCN can employ RK4 integration because its buffer remains fixed during the intermediate sub-steps of each timestep: only the final integrated state updates the buffer for the next step. This avoids the hidden-state drift that prevents the LSTM from using higher-order integrators. The TCN captures temporal patterns through learned convolutional filters rather than recurrent gating, offering an alternative that is more amenable to parallelization during training.

%--------------------------------------------------
\begin{table}[t]
\centering
\caption{Summary of the seven dynamics models evaluated in this work. LST and LTVL are the proposed PIVOT models; the remaining five serve as baselines. Physics-informed models leverage known dynamics equations, while data-driven models learn the full state-derivative mapping from data. Temporal models incorporate past state-action history.}
\label{tab:models}
\small
\begin{tabular}{@{}llcccc@{}}
\toprule
Category & Model & Dim. & Learned & Physics & Temporal \\
\midrule
\multirow{2}{*}{Fixed} & DST & 7 & -- & \checkmark & -- \\
 & DTVL & 5 & -- & \checkmark & -- \\
\midrule
\multirow{2}{*}{PIVOT} & LST & 7 & \checkmark & \checkmark & -- \\
 & LTVL & 5 & \checkmark & \checkmark & -- \\
\midrule
\multirow{3}{*}{Data-Driven} & MLP & 5 & \checkmark & -- & -- \\
 & LSTM & 5 & \checkmark & -- & \checkmark \\
 & TCN & 5 & \checkmark & -- & \checkmark \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]
\centering
\caption{Parameter counts for each learned model. All models share the same 858K-parameter backbone; the output layer scales with the number of predicted dynamics parameters. The LST model requires only 8 outputs, keeping total parameters near the backbone size, while the LSTM requires 19,141 outputs, inflating the total to 5.8M and increasing overfitting risk.}
\label{tab:param-counts}
\small
\begin{tabular}{@{}lrrr@{}}
\toprule
Model & Dynamics Params & Output Layer & Total \\
\midrule
LST & 8 & 2,056 & 859,912 \\
LTVL & 40 & 10,280 & 868,136 \\
MLP & 838 & 215,366 & 1,073,222 \\
TCN & 821 & 210,997 & 1,068,853 \\
LSTM & 19,141 & 4,919,237 & 5,777,093 \\
\bottomrule
\end{tabular}
\end{table}
